ptr-tidy: Automatic Rejuvenation of Raw Pointers in C++

### 2021-04-11 paper.tex
Add rest of paper from overleaf

\usepackage[numbers]{natbib}
\usepackage{xurl}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{abstract}
    \end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    Most ordinary computer users demand that the machine they are using provides them with a responsive, secure and productive environment to complete their tasks.
    The programs that are most responsible for this are complex systems programs such as the underlying operating system, device drivers and web browsers.
    Systems programming often involves memory management, that is requesting memory from the operating system to be used and managed by the program.
    This is done when the amount of memory that we need cannot be know at the time of compilation of the program.
    However, doing so creates the opportunity for memory safety errors~\cite{Dhurjati2003} which are notoriously challenging to avoid.
    If there are memory safety errors in a program, there is the possibility that it could be exploited by a malicious party to crash the program, expose sensitive data or gain remote code execution capabilities~\cite{Afek2007}.

    Historically systems programs are mostly written in C, C++ or a combination of the two and remain the dominant languages.
    They are lower level languages that involve manual memory management by the programmer.
    It is even more challenging to ensure that C and C++ programs contain no memory safety errors as evidenced by empirical evidence in industry, where in a presentation at the Linux Security Summit it is shown that in several popular projects such as Firefox, macOS, Ubuntu and Android all had over half of their CVEs attributed to issues with memory safety~\cite{Gaynor2020}.
    Similarly, in a presentation by Matt Miller, a security engineer at Microsoft, it is shown that around 70\% of their vulnerabilities that are addressed through a security updates are due to memory safety issues~\cite{Miller2019}. 
    This is at industry-leading companies who are renowned for hiring top talent and top developer education programs.

    However the continued use of C and C++, alongside industry mottos such as \emph{``Move fast and break things''}~\cite{Vardi2018} do not help by possibly diminishing the danger of memory safety error exploitation.
    However due to new regulations such as the GDPR and increasing oversight by regulators, any such exploitations which do expose sensitive data are more commonly exposed to the general public and punished by regulators.
    An example is a fine of £20m for a British Airways data breach by the British Information Commissioner's Office~\cite{ICO2020}.

    However simply not using these languages is not an option since their unmanaged nature makes them highly performant, and therefore the best option for systems such as a web browser.
    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers similar performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model also guarantee memory safety.
    The rewriting of a program in a new language, namely from C++ to Rust, is a colossal undertaking, especially given that Rust has a reputation for being difficult to learn.
    Instead, modern C++ guidelines encourage the use of \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create memory or resource handles which automatically eliminate entire classes of memory safety errors.
    Yet, we encounter further problems as C++ programs either do not actively follow this advice, or are \emph{legacy} programs that were written before these features were introduced to the language.

    We therefore propose a solution in the form of a tool, \emph{ptr-tidy}, to analyse and automatically refactor programs to follow the recommended use of smart pointers where the analysis can determine that such a change is syntactically correct and does not change the runtime behaviour of the program.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \section{Implementation}\label{sec:implementation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usepackage{minted}
\usepackage[numbers]{natbib}
\usepackage[font=bf]{caption}

\providecommand*{\listingautorefname}{Listing}

\setminted{frame=lines, autogobble, breaklines}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{abstract}
    Systems programming involves dynamically requesting memory that in the C++ language has to be manually managed by the programmer. 
    However, this leaves the opportunity for exploitable memory safety errors.
    Modern C++ recommends that smart pointer are used to automatically manage resources such as dynamic memory.
    However, many legacy programs may have been written before these features were introduces to the language.
    We developed the ptr-tidy tool in order to automatically rejuvenate such programs to use smart pointers.
    The tool uses the Clang and LLVM libraries in order to create a componentised and powerful design which could be reused by other refactoring or rejuvenation tools. 
    We present a case study and and apply the tool on a number of open-source C++ libraries to show that the tool can successfully rejuvenate non-trivial libraries and that the tool can be used on libraries from many areas of computer science. 
    We discuss in which circumstances the tool performs well and the current limitations.
    \end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    Most ordinary computer users demand that the machine they are using provides them with a responsive, secure and productive environment to complete their tasks.
    The programs that are most responsible for this are complex systems programs such as the underlying operating system, device drivers and web browsers.
    Systems programming involves memory management, that is dynamically requesting memory from the operating system to be used and managed by the program.
    This is done when the amount of memory that we need at runtime cannot be know at the time of compilation of the program.
    However, doing so creates the opportunity for memory safety errors~\cite{Dhurjati2003} which are notoriously challenging to avoid.
    A memory safety error can be one of:
    \begin{itemize}
        \item \textbf{Dangling pointers}, prematurely freeing an object that is still in use.
        \item \textbf{Double free}, deleting an object multiple times.
        \item \textbf{Never free}, never deleting an object.
    \end{itemize}
    There are other categories of memory safety errors such as buffer overflows which we do not consider in this paper.
    If there are memory safety errors in a program, there is the possibility that it could be exploited by a malicious party to crash the program, expose sensitive data or gain remote code execution capabilities~\cite{Afek2007}.

    Historically systems programs are mostly written in C, C++ or a combination of the two and these remain the dominant languages.
    They are lower level languages that involve manual memory management by the programmer.
    It is even more challenging to ensure that C and C++ programs contain no memory safety errors as evidenced by empirical evidence in industry, where in a presentation at the Linux Security Summit it is shown that in several popular projects such as Firefox, macOS, Ubuntu and Android all had over half of their CVEs attributed to issues with memory safety~\cite{Gaynor2020}.
    Similarly, in a presentation by Matt Miller, a security engineer at Microsoft, it is shown that around 70\% of their vulnerabilities that are addressed through a security updates are due to memory safety issues~\cite{Miller2019}. 
    This is at industry-leading companies who are renowned for hiring top talent and top developer education programs.

    % However the continued use of C and C++, alongside industry mottos such as \emph{``Move fast and break things''}~\cite{Vardi2018} do not help by possibly diminishing the danger of memory safety error exploitation.
    % However due to new regulations such as the GDPR and increasing oversight by regulators, any such exploitations which do expose sensitive data are more commonly exposed to the general public and punished by regulators.
    % An example is a fine of \pounds20m for a British Airways data breach by the British Information Commissioner's Office~\cite{ICO2020}.

    Unfortunately, simply not using these languages is not an option since their unmanaged nature makes them highly performant, and therefore the best option for systems such as a web browser.
    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers similar performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model also guarantee memory safety.
    The rewriting of a program in a new language, namely from C++ to Rust, is a colossal undertaking, especially given that Rust has a reputation for being difficult to learn~\cite{Astrauskas2019}.
    Instead, modern C++ guidelines encourage the use of \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create memory or resource handles which automatically eliminate entire classes of memory safety errors.
    Yet, we encounter further problems as C++ programs either do not actively follow this advice, or are \emph{legacy} programs that were written before these features were introduced to the language.

    We therefore propose a solution in the form of a tool, \emph{ptr-tidy}~\cite{Usov2021}, to analyse and automatically refactor programs to follow the recommended use of smart pointers where the analysis can determine that such a change is syntactically correct and does not change the runtime behaviour of the program.

    \begin{listing}
        \inputminted{c++}{code/leak.cpp}
        \inputminted{c++}{code/fix-leak.cpp}
        \caption{Example of a memory leak using manual management and how smart pointers correct this. Lines 4 and 5 show how interruptions in the program control flow can mistakenly cause memory leaks and how smart pointer correct this by automatically inserting all needed delete statements at compile time. All necessary headers are omitted for brevity}
        \label{lst:motivating-example}
    \end{listing}

    We provide a motivating example in \autoref{lst:motivating-example} where the first version of the function could have several control flow executions where the \texttt{Circle} object will leak.
    The example seems quite trivial, but can suddenly be quite non-trivial if the conditions in the \texttt{if} statement become calls into external libraries, which may or may not throw an exception.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    Our work closely relates to work done around the field of \emph{source code rejuvenation}, as defined by \citet{Pirkelbauer2010}. 
    We exactly aim replace outdated coding patterns with newer, higher-level abstractions, as well as fitting the notion that this tool would be applied once to a codebase, rather than be a reoccurring task.
    We will however use the terms rejuvenation and refactoring interchangeably in this paper as it is a term that most developers are familiar with.

    We believe research in this area is going to increase, as languages move and evolve much quicker than can be supported in enterprise environments, resulting in accumulating technical debt. 
    This is supported by the fact that other relevant work has been completed on this subject.

    First and foremost, we have evidence that this is a problem that exists in industry through a paper by \citet{Wright2013} which describes a tool used at Google for running such rejuvenations across a large codebase. 
    \citet{Wright2013} show an example of running a simple rejuvenation of upgrading an API call.
    They use Clang for the parsing of code into an AST and traverse the tree to identify and perform simple refactorings.
    Their flexible design allows them to implement many different simple refactorings using the same tool, rather than the focus we have for a single specific refactoring goal.

    Secondly, there has been separate academic research into rejuvenation tools such as the work by \citet{Huck2016} and \citet{Kumar2012}.
    The former also uses the Clang and AST driven approach like \citet{Wright2013}, while the latter uses a representation called IPR~\cite{Dos2011}, which is a general and efficient data structure for representing C++ programs.
    We have not evaluated IPR as a representation compared to the Clang AST, but we can however comment that whilst we appreciate the aim of the project to create a efficient and compiler-independent representation, the greater amount of documentation that Clang has, as well as it being the choice of several other successful projects leads us to believe that it may be the more suitable and mature option for any such research.

    There has also been industry interest in automatic rejuvenation tools between languages rather than within the same language. 
    This is done either to increase the use of a new, improved language by creating creating tools to aid transition, or in order to reuse existing compilers.
    For example, the Kotlin language team created the J2K~\footnote{\url{https://github.com/JetBrains/kotlin/tree/master/j2k}} transpiler in order to entice Java developers to switch to a language with improved language features such as null safety~\cite{Mateus2019}. 
    Similarly, several new languages such as Typescript and Dart have been recently developed which were designed not to be used by themselves, but to be transpiled to the popular Javascript language.
    These languages aim to solve shortcomings of the output language, whilst reusing the existing mature and efficient compilers or interpreters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    \subsection{Using Smart Pointers for Memory Management}\label{subsec:using-smart-pointers-for-memory-management}

    Generally pointers are used to give the program access to a resource that cannot be directly included in the program itself, such as allocated memory or a file.
    Pointers can also be used to pass large resources to functions without needing to create a copy, which would have a performance impact.

    Pointers can be created for objects both on \emph{the stack and the heap}~\cite{Stroustrup2000}. 
    The stack is an small memory store automatically managed for us by the compiler and is suitable for placing small, fixed sized parameters and arguments, such as perhaps loop counter variables.
    However, when objects must live beyond the lifetime of the current function call or if we do not know the size of an object until runtime, then we must storage these objects on the heap, occasionally called the \emph{free store}. 
    Unlike the stack, the memory on the heap must be explicitly managed by the programmer.

    However Stroustrup claims that pointers to objects allocated on the free store are dangerous and a \emph{plain old pointer}, or raw pointer as we will refer to them, should not be used to represent ownership~\cite{Stroustrup2018}.

    Instead we can use \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create resource handles which automatically eliminate resource leaks with no, or very little added overhead.
    RAII means that the management of a resource, namely its acquisition and release, is bound to its lifetime. 
    In the simplest case, this means that when a resource exists the scope of a function or method, it is released automatically.

    The first type of smart pointer are \emph{unique pointers} which have only a single unique owner at any time and therefore are ideal for resources which may not be copied, such as perhaps a global session dictionary.
    We can see how unique pointers and RAII work from out motivating example in  \autoref{lst:motivating-example}.
    Even though the \texttt{Circle} object is placed on the heap, the compiler can manage the smart pointer for us so that in all cases where the lifetime of the object ends, the object will be freed and ensures the resource will not leak. 
    The ownership of a unique pointer can be transferred between variables using \texttt{std::move} which moves the ownership from one variable to another, making the original reference invalid.
    % TODO insert example diagram possibly

    The second type of smart pointer are \emph{shared pointers} which are used for resources which may not necessarily have a single unique owner and so several owners \emph{share} ownership to the resource.
    Reference counting~\cite{Collins1960} is used to ensure that after the last owner loses access to the resource, it will finally be freed.
    Shared pointers are suitable to be used in most situations and such reference counted pointers are used by default for all values in the Python and Swift languages~\footnote{\url{https://docs.swift.org/swift-book/LanguageGuide/AutomaticReferenceCounting.html}}

    \subsection{Clang and LLVM}\label{subsec:clang-and-llvm}

    \begin{figure*}
        \centering
        \includegraphics[scale=0.75]{images/llvm.pdf}
        \caption{LLVM design diagram showing how language front-ends create LLVM IR that can be optimised and then used by an architecture back-end, such as x86 to create a machine code executable.}
        \label{fig:llvm-design}
    \end{figure*}

    As seen in \autoref{sec:background}, much similar research uses Clang and LLVM.
    LLVM~\cite{Lattner2004} is a compiler infrastructure project consisting of several compiler and toolchain technologies. 
    LLVM is centrally designed around a language independent intermediate representation (IR), which can be created by a number of language front-ends and is used by a number of architecture back-ends to generate machine code. 
    Program optimisations can then be applied by portable and reusable transformations on the IR.
    Any improvement in back-end code generation, as well as any new or improved IR optimisers can then benefit any of the supported language front-ends.
    It is therefore visible that this allows for a very modular and extensible design which can be seen in \autoref{fig:llvm-design}.

    Clang~\cite{Lattner2008} is such an LLVM front-end for C, C++ and a variety of other similar languages such as OpenGL.
    By using the LLVM infrastructure, it allows for faster and more efficient compilation of C++ compared to existing, older compilers such as the GNU Compiler Collection~\footnote{\url{https://gcc.gnu.org/}} (GCC)~\cite{Lattner2008}.
    However, most importantly for our research, the design of GCC makes it unsuitable for integration with other projects, such as analysis tools, whereas Clang allows access to several analysis tools as well as the Clang abstract syntax tree (AST) and parser internals via a standard API such as consumers and visitors~\cite{Duffy2014}.

    \begin{figure}
        \centering
        \inputminted{c++}{code/code-to-ast.cpp}
        \includegraphics{images/code-to-ast.pdf}
        \caption{Example of converting C++ code to Clang AST. All diagrams show equivalent representation of a small C++ code snippet.}
        \label{fig:code-to-ast-example}
    \end{figure}

    Further more, Clang also explicitly provides the \emph{LibTooling} C++ library for the creation of tools that leverage the Clang parsing front-end and AST.
    The Clang AST is different from ASTs produced by some other compilers in that it closely resembles both the written C++ code and the C++ standard~\footnote{\url{https://clang.llvm.org/docs/IntroductionToTheClangAST.html}}.
    For example, we can see in \autoref{fig:code-to-ast-example} that the AST stores much syntactic as well as semantic information in the form of source code locations and variable initialisation styles.
    The LLVM authors remark that this makes it very suitable for refactoring tools.

    The LLVM IR represents a language-independent, static single assignment (SSA)~\cite{Rosen1988} representation of the program.
    Variables in in SSA form can only be assigned once, which allows us to efficiently generate compact definition-usage graphs, which enable developers to simplify many dataflow optimizations and enables fast, flow-insensitive algorithms~\cite{Lattner2004}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \section{Methodology}\label{sec:methodology}

    \subsection{Parsing}

    The first capability our tool would need is to be able to parse C++ code.
    Doing this ourselves would be an undertaking that would take a significant amount of time in order to create a complete parser due to the size and complexity of the C++ language, therefore it was decided early to use an existing parser.
    The information in \autoref{subsec:clang-and-llvm} outlines why Clang is the most suitable parser to use, with the other benefit that it brings such as \emph{LibTooling}.

    \subsection{Analysis}\label{subsec:methodology-analysis}

    Next, our tool would need to be able to identify specifically which pointers are suitable for refactoring.
    An \emph{unique pointer} is a smart pointer object that represent a single unique owner of a region of memory at any time. 
    We will therefore need to develop a static analysis algorithm to identify the maximum number of owners a region of memory has throughout all points of a program. 
    If a memory region has at most a single unique owner at a time, we can refactor the initialising raw pointer to a unique pointer, otherwise the pointer can be refactored to a shared pointer.    
    This analysis will need to ensure that its result is correct, and will inevitably have to make some conservative assumptions about the program.

    \begin{figure}
        \centering
        \includegraphics[scale=1.0]{images/memory.pdf}
        \caption{Both the variables \texttt{a} and \texttt{c} end up pointing to memory address \texttt{0001}. Similarly variables \texttt{b} and \texttt{d} point to address \texttt{0003}. The memory at \texttt{0001} is used through two different pointers on line 5, meaning that it is not uniquely owned by a single owner. However, address \texttt{0003} does have a unique single owner at any time in the program, shown by the single incoming arrow to the memory region. Therefore the initialisation of variable \texttt{d} can be modelled as an ownership transfer from \texttt{b}}
        \label{fig:memory-pointers}
    \end{figure}

    A key intuition is that a region of memory, subsequent to its initialisation, is only able to gain additional owners if access to the region of memory is \emph{shared} to another variable, that is if a pointer to this memory is copied to a another pointer, at which point the original owner is longer the unique owner.
    If we copy the pointer to a global variable, the variable could suddenly be read by any other running threads in the program, so we have to take a conservative decision that the memory no longer has a single unique owner.
    If we copy the pointer to a local variable, then if the original variable remains in use, the memory will have two unique owners. 
    However if the original pointer is not used beyond being copied, this could actually be modelled as a \emph{transfer} of ownership, and the memory will still have a single unique owner at all points of the program. 
    We show this intuition in \autoref{fig:memory-pointers}, where the memory region \texttt{0001} is not suitable for being managed by a unique pointer, whilst region \texttt{0003} is.

    We can formalise our key intuition by introducing the concept of \emph{escape analysis}.
    Escape analysis is used to determine whether an object \emph{escapes} or is accessible from outside the method or thread that created the object~\cite{Choi1999}.
    It has been used in languages such as Java~\cite{Choi1999} to determine whether an object which is created within a method escapes.
    If it does not escape, meaning it is only used locally in the method, then it can be allocated on the stack rather than the heap as a performance optimisation.
    LLVM also uses escape analysis for a similar kind of optimisation~\footnote{\url{https://github.com/llvm/llvm-project/blob/main/llvm/lib/Analysis/CaptureTracking.cpp}}.

    It is now visible that the invariant of the unique pointer are only violated when:
    \begin{itemize}
        \item a pointer escapes, as then we do not know where or how the pointer may be used, so we must take a conservative approach and assume the memory region no longer has at most a single unique owner.
        \item a pointer is does not escape but access to the memory is shared rather than transferred to a local variable. 
    \end{itemize}

    \subsection{Analysis Target}\label{subsec:methodology-analysistarget}

    The existing escape analysis in LLVM is implemented to operate on LLVM IR code, as the IR representation, as mentioned in \autoref{subsec:clang-and-llvm}, is well suited for efficient analysis algorithms. 
    However, much of the existing work from \autoref{sec:related-work} which use Clang perform both their analyses and their source code refactoring on the Clang AST.
    We however propose an improved novel design which uses the Clang AST for parsing and refactoring and the LLVM IR for simple and efficient analysis.

    For the same input file, we can use the \emph{LibTooling} library to generate an AST and IR in-memory representation.
    Then, for any variable declaration (\emph{VarDecl}) in the Clang AST which allocates new memory on the heap, we locate this declaration in the IR.
    This is possible as we can instruct Clang to not discard variable names using a \texttt{fno-discard-value-names} flags, and as the combination of function signature and variable name in the translation unit (\emph{cpp file}) must be unique according to the One-definition rule~\cite{C++17}.

    \begin{listing}
        \inputminted{llvm}{code/code-to-ir.ll}
        \caption{The LLVM IR module representation of the C++ code in \autoref{fig:code-to-ast-example}. Listing has been summarised for clarity. The variable on line 5 can be matched to the variable \texttt{a} in \autoref{fig:code-to-ast-example} by matching the function signature and variable name.}
        \label{lst:code-to-ir}
    \end{listing}

    In \autoref{fig:code-to-ast-example}, given the variable declaration for the variable \texttt{a}, then as it is not a global variable, we can walk its parent nodes until we reach a \emph{FunctionDecl}.
    We are now able to use the function name to find the corresponding function declaration in the IR module, which can be seen in  \autoref{lst:code-to-ir}.
    Note that although the \texttt{main} function name corresponds exactly, other function names in the IR module will have mangled names~\footnote{\url{https://www.ibm.com/support/knowledgecenter/en/ssw_ibm_i_72/rzarg/name_mangling.html}}.
    We again use Clang library functions to mangle function names if necessary so they can be found in the IR. 
    Given a function in the IR, we can use its symbol table to use the declaration name to find the IR value on which we will perform our analysis.

    \subsection{Choice of Escape Analysis Algorithm}

    The initial choice of escape analysis algorithm is the existing LLVM algorithm mentioned in \autoref{subsec:methodology-analysis}.
    As it is already developed for use on LLVM IR, it is therefore trivial to integrate into the tool which allows us to quickly create a minimal viable product. 
    This allowed for rapid successful prototyping of the tool and therefore reduces risk of the chosen design being unsuitable by following rapid development processes~\cite{Martin1991}.
    We show in \autoref{subsec:methodology-systemdesign} that the design of our tool means that the initial choice of algorithm will not place a limit on the final accuracy or power of the tool.

    Investigation of the algorithm shows that it is a conservative analysis with some limits on the analysis complexity so that it does not greatly increase Clang compilation time.
    The analysis uses the definition-usage graph of the value being analysed to add all the instructions that use the value to a work queue.
    Each item in the work queue is then processed and the algorithm dispatches on the type (\emph{opcode}) of the IR instruction.
    It is then checked whether each instruction within the context that it is being used in could cause the value to escape.
    The key limitation that we see of this algorithm is that it is a intraprocedural analysis.
    Pointers are very often used across function boundaries and the conservative decisions the algorithm will take in these situations might limit the effectiveness of the tool in identifying all possible refactoring opportunities.

    \subsection{Source code rewriting}

    The analysis gives a result of whether the IR value escapes, a boolean \texttt{true} or \texttt{false}.
    For a \texttt{false} result, the value declaration is refactored into a unique pointer, otherwise a shared pointer.
    However, the rewriting is identical for both cases with just the string tokens that we replace for the original type differing.
    The rewriting was developed using a rule based approach, using test driven development (TDD) as this allowed rapid but correct development by continuously being able to check that additional created rules still cause all the tests to pass, causing no regressions. 
    The five cases of code (\emph{or rewriting rules}) that needs to be rewritten are:
    \begin{itemize}
        \item Declaration types.
        \item Variable initialisation and assignment expressions.
        \item Pointer delete statements.
        \item Function return types.
        \item Pointer Usages.
    \end{itemize}

    Rewriting declaration types are the easiest of all. We can directly replace the type of the variable declaration node that we already have access to. 
    For example for \mintinline{c++}{int *ptr}, we can extract the type of the variable, and format it into the string \mintinline{c++}{"std::shared_ptr<{}>"} or \mintinline{c++}{"std::unique_ptr<{}>"}, where the \texttt{\{\}} represent the replacement field.

    Initialisation and assignment expressions are less simple due to two reason.
    First of all, the initialisation of a variable does not necessarily have to occur immediately at the point of declaration.
    Therefore we need to find all the points of initialisation or value assignment to a variable in the AST.
    Secondly, there are many different ways to initialise a variable~\cite[p.~196]{C++17}, with the most common initialisation methods using:
    \begin{itemize}
        \item Braced-lists \mintinline{c++}{std::string s{}}.
        \item Equal sign \mintinline{c++}{std::string s = "hello"}
        \item parentheses expression list \mintinline{c++}{std::string s("hello")} 
    \end{itemize}
    For pointer variables, this is convoluted even more depending on if the pointer gets initialised using an existing pointer \mintinline{c++}{int* a = nullptr; int *b = a;}, or if it gets initialised using a new statement \mintinline{c++}{auto p = new Circle{Point{0,0},10}}.
    In the case of a new statement, we can initialise the rewritten smart pointer using a value initialiser \mintinline{c++}{std::make_unique<>()} or the shared pointer equivalent. 
    Otherwise, we can create a smart pointer from a pointer value using by the default equal sign copy initialiser or parentheses expression list.

    Rewriting pointer deletion statements also require us to traverse the AST to find \mintinline{c++}{delete} statements for the variable declaration node that we are rewriting.
    However, they do not require any complicated logic, and we can upon finding relevant statements remove them from the source code.

    If the variable declaration is a variable that gets returned at the end of the function, then the return type of the function signature will need to be rewritten too. 
    We explicitly check for each variable declaration that we handle if it gets returned, and if so, we can rewrite the function return type the same way that we handle variable declaration types.

    Lastly, smart pointer have overloaded operators for the arrow \texttt{->} and dereference \texttt{*} operators so that they behave identically as if the value was a raw pointer.
    This simplifies our task as it reduces the amount of rewriting we have to do for each usage of a specific pointer value.
    However, there is no such overloaded operator for passing smart pointer as function arguments. 
    In this case, the compiler will find a type error, when it might expect a type such as \mintinline{c++}{int *}, but actually receives \mintinline{c++}{std::unique_ptr<int>}. 
    Therefore, we again have to traverse the AST to find all function call expressions which involve the variable declaration that we are currently handling, and we need to add \mintinline{c++}{get()} calls to the smart pointer values to get a raw pointer to the managed objects.

    We use the Clang \emph{Rewriter}\footnote{\url{https://clang.llvm.org/doxygen/classclang_1_1Rewriter.html}} class as a high level abstraction compared to handling files and text buffers ourselves, which would likely be error-prone for buffer overflows.

    \begin{figure}
        \centering
        \begin{minted}{c++}
        char *test() {
            char* a = new char[2];
            a[0] = 'a';
            printf("\%s", a);
            return a;
        }
        \end{minted}

        \includegraphics{images/arrow.pdf}

        \begin{minted}{c++}
        std::shared_ptr<int> test() {
            std::shared_ptr<char> a = std::shared_ptr<char>(new char[2]);
            a.get()[0] = 'a';
            printf("\%s", a.get());
            return a;
        }
        \end{minted}

        \caption{An example of all the rewriting rules except pointer delete statements. This example solely shows rewriting and does not necessarily follow respect any smart pointer viability analyses. We see how both the declaration and equal sign copy initialisation of \texttt{a} have been rewritten to into a \texttt{shared\_ptr}. All the uses of \texttt{a}  that require a raw pointer have a \texttt{get()} call added to obtain the managed memory.}
        \label{fig:rewriting}
    \end{figure}

    In \autoref{fig:rewriting}, we show an example that incorporates all the cases in which code needs to be rewritten except pointer delete statements.
    It is visible how in the majority of cases, the first two rules will be applicable as every new smart pointer need to have the correct type and be initialised to the right type of object. 
    The fifth rule will also likely be applicable in most cases as variables tend to not be unused and variables which were pointers will be used in functions expecting pointers, as they otherwise could be stack variables.

    \subsection{System Design}\label{subsec:methodology-systemdesign}

    \begin{figure*}
        \centering
        \includegraphics[scale=0.65]{images/design.pdf}
        \caption{System design diagram of the ptr-tidy tool.}
        \label{fig:system-design}
    \end{figure*}

    We show the full system design diagram in \autoref{fig:system-design}, showing the full sequence of operations that are executed in order to transform and refactor the input file containing raw pointers into the output file that uses smart pointers.
    The input file is parsed into a Clang AST, which is traversed to find variable declarations.
    The corresponding variable declaration is found in the IR, and used as input to the analysis algorithm.
    The output of the analysis is used in the rewriting component, after which all the changes made in rewriting the program are consolidated and output as the new, improved program.

    We believe this design is noteworthy as it closely follows the system design of the general LLVM infrastructure in \autoref{fig:llvm-design}, by having a Clang parser front-end, an analysis middle-end, and a rewriter back-end.
    We also roughly follow the pipeline architectural design, as described by \citet{Schmidt2013}, by having a series of transformation steps, which each have an input and one or more outputs.
    A series of transformation steps are composed until we obtain our end output file.
    This allows us to conform to the Single-responsibility principle as proposed by \citet{Martin2002}.
    This gives us a highly flexible design where different analysis algorithms can be very easily added and trialled for the analysis component, meaning that potentially more powerful analyses can be tested without needing any changed to the parsing or rewriting components. 
    Similarly, given that our analysis is executed on the language-agnostic LLVM IR, the analysis component could be reused by rejuvenation tools for any other languages, allowing our work to be used in future research.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    Through our research, we would like to answer the question of whether our implemented tool can truly improve C++ code by seeing if the tool can be used outside small examples on larger existing programs, as well as if the refactored results produce code which can be compiled and can be judged to be better.
    Secondly, if the tool is able to do this, we want to evaluate how well the tool performs by analysing the effectiveness of the analysis algorithms.

    In order to evaluate the objectives above, hypotheses were made:

    \begin{itemize}
        \item \textbf{Hypothesis 1} The automatic translation of raw pointers into smart pointers can be used to make existing C++ code more modern, safe and extendable.
        \item \textbf{Hypothesis 2} The tool can reliably identify situations in which unique pointers should be used.
    \end{itemize}

    Each of these hypotheses will require experimentation to be verified.
    We will answer hypothesis 1 through a case study and we will answer hypothesis 2 through the gathering of relevant metrics on several more projects.
    This section will detail the justification, process, and results for these experiments.

    We run our experiments on a Windows 10 Pro x64 machine using the Windows Subsystem for Linux, with an AMD Ryzen 5 PRO 4650U and 16 GiB RAM, using TinyXML-2 v8.0.0.
    The ptr-tidy tool was compiled with GCC v10.2.0 using CMake release profile, using Clang and LLVM v11.1.0 libraries.

    \subsection{TinyXML-2 Case Study}\label{subsec:case-study}

    In order to answer hypothesis 1, it was decided to run a case study instead of applying the tool in large on a range of open source projects. 
    This was decided as an analysis into a well chosen target project will be able to show relevant success and failure cases that we can extrapolate onto more general C++ projects of the same kind. 
    If the impact of the success cases outweighs the the faults of the failure cases, then we will be able to judge the our hypothesis as being true.

    We decided on using the \emph{TinyXML-2} library as our chosen case study~\footnote{\url{https://github.com/leethomason/tinyxml2}}.
    TinyXML-2 is a C++ XML parsing library that can parse an XML document and build from that a Document Object Model that can be read, modified, and saved.
    TinyXML-2 has over three thousand \emph{GitHub stars} which puts it in the top 1\% of projects on the hosting website, which can be used as a rough metric for its popularity and good code quality~\cite{Sanatinia2016}.
    It is also for example used in the \emph{libigl}~\cite{Libigl} geometry processing library,  which itself is used by a variety of large industry companies.
    It is released under a \emph{zlib} license~\footnote{\url{https://opensource.org/licenses/Zlib}} which allows us to use and alter the software.
    What makes library a particularly attractive choice is the simplicity of the library.
    The library does not make use of exceptions, RTTI or the STL~\cite{C++17}, and therefore most importantly does not make use of smart pointers. 
    The library is also around three thousand lines of code in length, making it large enough to likely have a variety of edge cases, whilst remaining small enough for manual inspection.

    \begin{listing}
        \inputminted{cpp}{code/tinyxml2-example.cpp}
        \caption{Example of a method in the TinyXML-2 library which could be refactored to use smart pointers as it returns an owning pointer to a newly created object.}
        \label{lst:tinyxml-refactor-example}
    \end{listing}

    In particular, what makes this suitable for the target of our case study over perhaps other similarly small and simple libraries is that due to parsing varying documents at runtime, the library is particularly heavy in pointer logic and memory allocations.
    We can in \autoref{lst:tinyxml-refactor-example} an example of a method from the library which would be well suited for refactoring to use smart pointers.
    The heavy use of pointers and memory allocation is what leads us to believe that the case study into this library will generalise well onto all other C++ projects.

    \subsubsection{Experiment}
